Flat-histogram method comparison on the 2D Ising model

### 2020-06-29 ising/referee-response.md

Dear editor,

We appreciate the referees' careful reading of our paper, and feel that their
feedback has led to a significant improvement. We have addressed all of the
first referee's comments by added two paragraphs to the introduction and a
section in the paper to 'Address the general problem of all "flat-histogram"
methods'. We also strengthen the abstract to address the concerns raised by the
second referee. We point out that: (i) Our paper is the first of its kind (to
the best of our knowledge) to compare ``production run'' WL with other recent
flat-histogram methods. (ii) It is important to test SAD's convergence
properties on the 2D Ising where the range of energies is easily known.

@@

The first referee points out that 'Simulation set-up and analysis protocols are
sound and give confidence in the quality of the presented data and
conclusions.'

The referee also asks that three additions be made to the paper.

(i) "It would be worthwhile to work out the [slower convergence for SAD due to
the energy range being an a priori parameter in WL/MUCA] in the paper, and, if
possible, quantify the loss in efficiency coming from energy range
determination."

We have added a paragraph in the discussion section that explains the loss in
efficiency SAD suffers by not specifying an energy range a priori. We are able
to quantify this by identifying the last time SAD finds a new important energy
and then finding the fraction of moves (in the end) that are outside the range
of interesting energies. We find that SAD histogram counts outside the interesting
energy range are negligible. We mention this in the discussion paragraph.

@@

We believe that this paper helps by bridging the disconnect between *users* of
WL and developers of improved flat-histogram methods. The majority of
developers use pure WL when comparing multiple flat-histogram methods since
there is no direct comparison of WL followed by a production run with methods
such as 1/t-WL or SAMC.

(ii) "It is not clear[ly] understood why the current calculations were not
included in the original article. Although the results presented by the
[authors] are original, they are only a simple extension of the first work, and
do not meet the standards of PRE readers."

The second referee asks an important question here to which there are two
answers. We only became aware of "production run" WL as a result of the
response to our first paper. As mentioned (hinted at) in the acknowledgement,
Johannes Zierenberg emailed us after reading our paper and suggested that we
test WL followed by a production run with a number of flat-histogram methods.
Due to the lack of literature detailing its performance vs other flat-histogram
methods, the 2D Ising model seems the logical starting point for such a
comparison.

Second, SAD was initially tested on methods where it could benefit from knowing
the temperature range of interest in advance. It is critical to test it's
convergence properties on a different type of system where the range of
energies is easily known (we find that this does not measurably bias in favor
of WL methods). Also in the case of the 128 X 128 2D Ising system, the
computation time is quite lengthy (more than 6 months for some calculations)
representing a superior and robust comparison result.

In response to the second referee's comments, we have better highlighted these
important points in the abstract, discussion paragraph, and conclusion.

### 2020-06-29 ising/ising-body.tex
made ising paper edits and added a few more citations on replica exchange.

% optimized ensemble paragraph
Nadler \emph{et al.} and Trebst \emph{et al.} systematically examined optimized
ensembles due to performance issues flat-histogram methods faced when
confronted with hidden energy barriers~\cite{nadler2007generalized,
nadler2007dynamics, trebst2004optimizing}. Trebst \emph{et al.} used their
optimized ensemble to systematically set the temperature range(s).

% replica exchange paragraph
Replica exchange Wang-Landau (REWL) was introduced~\cite{vogel2013rewl}.
Replica-exchange Wang-Landau has been developed with great
success~\cite{perera2016magnetic} and adapting flat-histogram methods to use
such an optimized ensemble is doable

@@

Figure~\ref{fig:N128-cv-error} shows that the SAD algorithm converges
signiticantly more slowly that the converging WL methods.
%We interpret this as due to the
%fact that the SAD simulation must take the time to discover the energy range
%of interest dynamically.
The SAD algorithm on average takes around
\davidsays{$10^9$} moves to identify as important all \davidsays{8192}
negative energy states, which is around \davidsays{twice as long} as
the number of moves that
the WL methods require in order to explore all the states in this
energy range.  The main difference between the convergence of these
methods is that the WL-based methods decrease $\gamma_t$ far more rapidly,
which leads to more rapid convergence.  The Wang-Landau approaches can
get away with this because the range of energies of interest is given
as an input rather than an output, allowing a more aggressive schedule
of reduction of $\gamma_t$.  This aggressive behavior is precisely what
requires that Wang-Landau methods be followed by some correction stage
(either $1/t$ or a production run) in order to correct residual errors.

%% The two WL simulations followed by a production run highlight that
%% choosing too large a $\gamma^{\text{WL}}_t$ will result in poor convergence
%% properties.

\section{Conclusion}
We find that SAD, $1/t$-WL, and WL followed by a production run (with an
adequately small $\gamma^{\min}$) demonstrate excellent and robust
convergence. They all converge more rapidly than SAMC, and unlike pure
WL do not suffer from error saturation. We find that for larger Ising
systems SAD reduces the update factor more slowly (and conservatively)
than $1/t$-WL and WL followed by a production run. This means that SAD
will take proportionately more moves to converge to the same value as
$1/t$-WL as system size is increased.  However, the WL methods are
solving a simpler problem because they are given the energy range
\emph{a priori}, rather than a temperature range of interest such as
SAD requires.  The 2D Ising model is a system where the energy range
is known \emph{a priori.}, which makes a comparison between WL methods
and SAD more biased in favor of WL methods. For the general case in
which the energy range of the system is not known and where a range of
desired temperatures is known, this will make the SAD method
considerably more convenient, and quite possibly more efficient than a
process involving multiple simulations to determine the energy range
of interest. This work also dramatically demonstrates that WL followed
by a production run performs extremely well and is preferable to pure
WL for ensuring both ergodicity and detailed balance. While this has long
been thought to be the case and that a multi-canonical run where WL is used to
determine the parameters is the only way WL should be used, this research
represents the first conclusive comparison among all of these
flat-histogram methods.

>>>>

% optimized ensemble paragraph
A problem that flat-histogram methods face is that the convergence rate is
impacted by energy barriers and bottleneck which make traversing the phase
space difficult. Nadler \emph{et al.} and Trebst \emph{et al.} systematically
examined optimized ensembles due to performance issues flat-histogram methods
faced when confronted with hidden energy barriers~\cite{nadler2007generalized,
nadler2007dynamics, trebst2004optimizing}. The goal of these methods is to make
the statistical errors uniform such that low energy errors do not dominate over
high energy errors.

% replica exchange paragraph

Parallel tempering (replica exchange) is a powerful method for overcoming
energy barriers~\cite{geyer1995annealing, hukushima1996exchange,
hansmann1997parallel}. The method works by breaking the simulation up into many
`replicas' and simulating in parallel. At certain times, neighboring replicas
are allowed to exchange which enables the phase space to be more evenly
explored.

Flat-histogram methods, such as Replica exchange Wang-Landau
(REWL)~\cite{vogel2013rewl,perera2016magnetic}, have been proposed that combine
histogram and replica exchange methods. This demonstrates that flat-histogram
methods can be extended to non-flat-histogram methods.

@@

Figure~\ref{fig:N128-cv-error} shows that the SAD algorithm converges
significantly more slowly than the converging WL methods.
The SAD algorithm on average takes around
$10^9$ moves to identify as important all 8192
negative energy states, which is around twice as long as
the number of moves that
the WL methods require in order to explore all the states in this
energy range.  The main difference between the convergence of these
methods is that the WL-based methods decrease $\gamma_t$ far more rapidly,
which leads to more rapid convergence.  The Wang-Landau approaches can
get away with this because the range of energies of interest is given
as an input rather than an output, allowing a more aggressive schedule
of reduction of $\gamma_t$.  This aggressive behavior is precisely what
requires that Wang-Landau methods be followed by some correction stage
(either $1/t$ or a production run) in order to correct residual errors.

\section{Conclusion}
We find that SAD, $1/t$-WL, and WL followed by a production run (with an
adequately small $\gamma^{\min}$) demonstrate excellent and robust
convergence. They all converge more rapidly than SAMC, and unlike pure
WL do not suffer from error saturation. We find that for larger Ising
systems SAD reduces the update factor more slowly (and conservatively)
than $1/t$-WL and WL followed by a production run. This means that SAD
will take proportionately more moves to converge to the same value as
$1/t$-WL as system size is increased.  While the WL methods are
are given the energy range
\emph{a priori}, rather than a temperature range of interest such as
SAD requires, we find that the SAD histogram counts for energies outside the range of interest are negligible. For the general case in
which the energy range of the system is not known and where a range of
desired temperatures is known, the SAD method is
considerably more convenient, and quite possibly more efficient than a
process involving multiple simulations to determine an energy range
of interest. This work also dramatically demonstrates that WL followed
by a production run performs extremely well and is preferable to pure
WL for ensuring both ergodicity and detailed balance. While this has long
been thought to be the case and that a multi-canonical run where WL is used to
determine the parameters is the only way WL should be used, this research
represents the first detailed comparison among all of these
flat-histogram methods.
